The Nuclear Norm and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-22 conf_article/neon.tex
Ky Fan k-norm introduced

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam and AdamW \cite{kingma2014adam}, \cite{loshchilov2017decoupled}, recently proposed Muon has shown promising results on training very large models \cite{jordan2024muon,ahn2025dioncommunicationefficientoptimizerlarge}. Its key difference from earlier algorthims is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

@@

\section{Related Work}
What to include: Muon and its upgrades (say that the same can be done with Neon), alternatives to Newton-Schulz, nuSAM and so on.

@@

\subsection{Norms $\normfstar{\cdot}^\dagger$ and $\normftwo{\cdot}^\dagger$}
Here we provide the derivation of these norms and plot them in the case of $2\times2$ matrices.
\subsection{Updates derivations}

Derivation of \cref{eq:muon_update} follows from \cref{eq:nsgd_muon_update} with $\alpha=1$. Indeed, $\norms{\cdot}^\dagger=1\cdot\normn{\cdot} + 0 \cdot \normf{\cdot} $.

Derivation of \cref{eq:nsgd_muon_update}:
Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.

Let us note that $\Delta = \alpha (\mU \mV^\top) + (1-\alpha) \frac{\mM^k}{\normf{\mM^k}}$ delivers this value. Indeed, by the trace property, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \alpha \mU \mV^\top + (1-\alpha) \frac{\mU \mSigma \mV^\top}{\normf{\mM^k}}> = \alpha \tr \mSigma + (1-\alpha) \normf{\mM^k} = \normfstar{\mM^k}$, which completes the proof.

Derivation of \cref{eq:neon_update} follows from \cref{eq:nsgd_neon_update} with $\alpha=1$. Indeed, $\normn{\cdot}^\dagger=1\cdot\norms{\cdot} + 0 \cdot \normf{\cdot} $.

Derivation of \cref{eq:nsgd_neon_update}:
Since $\norm{\cdot}^\dagger = \normftwo{\cdot}^{\dagger\dagger} = \normftwo{\cdot}$, the goal is to reach $\normftwo{\mM^k} = \alpha \sigma_1 + (1-\alpha) \normf{\mM^k}$.

@@

>>>>

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_\mathrm{KF-k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam and AdamW \cite{kingma2014adam}, \cite{loshchilov2017decoupled}, recently proposed Muon has shown promising results on training very large models \cite{jordan2024muon,liu2025muon}. Its key difference from earlier algorthims is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

    {\bf $\normkfk{\mM^k}^\dagger$ and Muon, Neon, and Dion without error feedback}
        We remind the reader that the Ky Fan k-norm \cite{fan1951maximum}, which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proven that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see Appendix). Special cases of Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

        \begin{lemma}\label{lemma:ky_fan_update}    
            When $\norm{\cdot} = \normkfk{\mM^k}^\dagger$, \cref{eq:our_update} turns into:

            \begin{equation}\label{eq:ky_fan_update}
                \mX^{k+1} = \mX^{k} - \eta \sum_{i=1}^{k}u_i v_i^\top
            \end{equation}
        \end{lemma}

        This lemma recovers the updates of Muon, when $k=\min\{m, n\}$, of Neon, when $k=1$, and of Dion \cite{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback.

        Moreover, one can consider F-KF-k-norm = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback.

@@

\section{Related Work}
As Muon \cite{jordan2024muon} is a very successful and popular optimizer for functions of weight matrices, a lot of reseach has been put into explaining its success and further improving its performance.

Regarding the second point, in less than a year, a large number of applications and improvements of Muon has been proposed. \cite{liu2025muon} adapted the algorthm for training language models larger than NanoGPT. \cite{shah2025practical} organized efficient hyperparameter transfer by combining Muon with maximal update parametrization. \cite{chen2025cosmoshybridadaptive} applied computationally intensive updates of SOAP optimizer to a low-dimensional "leading eigensubspace" while using memory-efficient methods like Muon for the remaining parameters. \cite{amsel2025polar} proposed a more efficient alternative to Newton-Schulz operations. \cite{si2025adamuon} introduced AdaMuon which combines element-wise adaptivity with orthogonal updates.

Regarding the first point, simplistic derivation of \cite{bernstein2025deriving} which is based on \cite{bernstein2024oldoptimizernewnorm} excluded, there has been a gap in theory behind Muon, which, as it seems to us, is not even now completely closed. \cite{kovalev2025understanding} has provided convergence guarantees of Muon in various settings, from which, however, Muon's supremacy cannot be recovered. Indeed, although the obtained bounds depend on the norm choice, the asymptotics of the convergence remain the same as for NSGD and other optimizers, $K = \cO(\epsilon^{-4})$ in a L-smooth stochastic case.

Similar drawback has a recent article \cite{riabinin2025gluon}, where L-smoothness assumption is replaced with a more practical $(L_0, L_1)$-smoothness. The authors derived from their theorems optimal stepsizes for Muon and Scion that match fine-tuned stepsizes reported by \cite{pethick2025training}. But still, they did not explain why, for instance, NSGD is inferior to Muon in training large-language models.

@@

\subsection{Norms $\normfstar{\cdot}^\dagger$, $\normftwo{\cdot}^\dagger$, and $\normkfk{\cdot}^\dagger$}
Here we provide the derivation of these norms and plot them in the case of $2\times2$ matrices.

\subsection{Updates derivations}

Proof of \cref{lemma:muon_update} follows from \cref{eq:nsgd_muon_update} with $\alpha=1$. Indeed, $\norms{\cdot}^\dagger=1\cdot\normn{\cdot} + 0 \cdot \normf{\cdot} $.

Proof of \cref{lemma:nsgd_muon_update}:
Since $\norm{\cdot}^\dagger = \normfstar{\cdot}^{\dagger\dagger} = \normfstar{\cdot}$, the goal is to reach $\normfstar{\mM^k} = \alpha \tr{\mSigma} + (1-\alpha) \normf{\mM^k}$.

Let us note that $\Delta = \alpha \mU \mV^\top + (1-\alpha) \frac{\mM^k}{\normf{\mM^k}}$ delivers this value. Indeed, by the trace property, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \alpha \mU \mV^\top + (1-\alpha) \frac{\mU \mSigma \mV^\top}{\normf{\mM^k}}> = \alpha \tr \mSigma + (1-\alpha) \normf{\mM^k} = \normfstar{\mM^k}$, which completes the proof.

Proof of \cref{lemma:neon_update} follows from \cref{eq:nsgd_neon_update} with $\alpha=1$. Indeed, $\normn{\cdot}^\dagger=1\cdot\norms{\cdot} + 0 \cdot \normf{\cdot} $.

Proof of \cref{lemma:nsgd_neon_update}:
Since $\norm{\cdot}^\dagger = \normftwo{\cdot}^{\dagger\dagger} = \normftwo{\cdot}$, the goal is to reach $\normftwo{\mM^k} = \alpha \sigma_1 + (1-\alpha) \normf{\mM^k}$.

@@

Proof of \cref{lemma:ky_fan_update}:
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal to reach $\normkfk{\mM^k}$.

Let us note that $\Delta = \sum_{i=1}^{k} u_i v_i^\top$ delivers the value. Indeed, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} u_i v_i^\top> = \sum_{i,j=1}^{r, k}\<u_i \sigma_i v_i^\top, u_j v_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^k}$, which completes the proof.
